Embedding Intuitionistic into Classical Logic

### 2022-07-14 CSL/lipics-v2021-EIICL.tex
Overview

Most state-of-the-art first-order theorem provers use classical logic~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}. Besides historic reasons a main advantage of classical logic over intuitionistic logic is the existence of convenient normal forms which enable reasoning techniques such as superposition and resolution. However there are also some drawbacks to this approach, in particular the non-constructiveness of classical logic poses a problem in synthesis-related tasks.

Several approaches for program extraction from classical proofs exist in literature. The first is simply to examine under which circumstances classical validity entails intuitionistic validity and therefore constructiveness. Most famously this holds for $\Pi_2$-formulas in Peano Arithmetic~\cite{friedman1978classically}. There are also other circumstances under which this entailment holds~\cite{schwichtenberg}, however this approach cannot encapsulate the general case. Another popular approach is to equip classical proofs with operational semantics~\cite{Control1, Parigot1}. In this case we necessarily lose realizability at least of $\exists$ and $\vee$, which - one could argue - is why we want a program in the first place.

Intuitionistic logic~\cite{sep-logic-intuitionistic} is a non-classical logic in which the law of excluded middle ($A\vee \neg A$) is not a theorem. A key motivation for studying intuitionistic logic is the Curry-Howard correspondence~\cite{howard1980formulae} which allows an effective bi-interpretation of proofs and programs.

One aspect that makes intuitionistic validity more difficult than classical validity is that it is not dual to satisfiability, i.e. unsatisfiability of $\neg\varphi$ is not sufficient for validity of $\varphi$. In fact intuitionistic satisfiability is equivalent to classical satisfiability and thus \verb+NP+-complete, while validity for intuitionistic logic is \verb+PSPACE+-complete~\cite{statman1979intuitionistic}. In classical logic this duality is the foundation of methods like resolution, tableaux~\cite{fitting2012first} and forms the backbone of most automated theorem provers. There is however a obvious dual notion to intuitionistic validity in the existence of Kripke counter-models, a standard semantics for intuitionistic logic~\cite{sep-logic-intuitionistic}. We will show that for each formula $\varphi$, if there exists a counter-model then there is one with fixed geometry completely determined by the structure of $\varphi$. This key observation will allow us to give a complete and effective embedding of intuitionistic logic into classical logic, i.e. for each formula $\varphi$ we give a formula $\varphi^\#$ that is classically valid if and only if $\varphi$ is intuitionistically valid, as well as an effective procedure to translate counter-models of the formulas. Intuitionistic validity can then be established using a classical prover by
\begin{itemize}
	\item performing our translation on the input formula $\varphi$ to obtain $\varphi^\#$
	\item checking the classical validity of $\varphi^\#$ with the prover resulting in a proof or counter-model
	\item translating the proof/counter-model of $\varphi^\#$ to a (intuitionistic) proof/counter-model of $\varphi$.
\end{itemize}

>>>>

Constructive mathematics refers to a flavour of mathematics in which existence of an object can only be established by explicit construction, as opposed to classical mathematics where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction. Besides the philosophical issues that one may take with classical mathematics, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation when looking from the position of computer science in that intuitionistic proofs correspond directly to computer programs, which is expressed in the Curry-Howard-Isomorphism~\cite{howard1980formulae}. The formalism usually associated with constructive mathematics is intuitionistic logic, which essentially differentiates itself from classical logic by the facts that the law of excluded middle $A\vee\neg A$ and double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.

However comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and ILTP~\cite{iltp}, we see that classical provers currently have a far better success rate. Furthermore all the most popular and used first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover} utilize classical logic. This difference can be partially explained by fundamental differences between the logics. First of all in the propositional case determining intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas the analogous decision problem for classical logic is only \verb+coNP+-complete~\cite{cook1971complexity}. A further advantage of classical logic is the existence of calculi that are particularly suitable for automation like superposition~\cite{bachmair2001resolution} which rely on the existence of convenient normal forms such as CNF and the duality between validity and satisfiability, i.e. to show validity they show unsatisfiability of the negation. While some kind of normal form would of course obtainable in the intuitionistic case, crucially the duality does not hold, e.g. $\neg(A\vee \neg A)$ is not satisfiable while $A\vee\neg A$ is not a theorem. Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly constructing a cut-free proof, enhanced by search strategies such as focussing and polarization. This strategy generally leads to a much more complex search and is difficult to implement efficiently. Finally in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, this is particularly true for the propositional case, i.e. SAT-solvers.

With this work we want to open the path for leveraging the progress that has been made in classical theorem proving for intuitionistic logic. To that end we give an embedding from intuitionistic logic into classical logic. That is we give an effective procedure that gives for each formula $\varphi$ we a modified formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. Interestingly the reverse direction, the famous double negation translation, is a long established translation that goes back to Glivenko~\cite{glivenko1929quelques} in the propositional case and G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit} in the first-order case. In the propositional case it is particularly simple: $\varphi$ is classically valid if and only if $\neg\neg\varphi$ is intuitionistically valid. Intuitively the translation collapses for each subformula $\psi$ the truth values of $\psi$ and $\neg\neg\psi$, which are classically but not intuitionistically equivalent. This gives us a first idea why the reverse direction is perhaps more difficult: We need to expand the truth values of $\psi$ and $\neg\neg\psi$, i.e. if they both occur in $\varphi$ we must have a way to (classically) assign different truth values to their respective counterparts in $\varphi^\#$. In particular this necessitates the introduction of new propositional variables, which already marks a big difference to the double negation translation.

In addition to the translation of formulas we give an effective translation of counter-models, i.e. for each intuitionistic counter-model for $\varphi$ we give a classical counter-model for $\varphi^\#$ and vice versa. In fact the existence of counter-models is a key notion. For classical logic it just corresponds to the satisfiability of the negation. However for intuitionistic logic it forms a proper dual to validity, while satisfiability of the negation is not necessary for invalidity and would therefore not form a proper dual. Transforming and reducing counter-models is what ultimately enables our translation.

Finally we should mention that approaches to obtain constructive content from classical proofs exist in literature, however they are insufficient in our view. The first is simply to examine under which circumstances classical validity entails intuitionistic validity and therefore constructiveness. Most famously this holds for $\Pi_2$-formulas in Peano Arithmetic~\cite{friedman1978classically}. There are also other circumstances under which this entailment holds~\cite{schwichtenberg}, however this approach cannot encapsulate the general case. Another popular approach is to equip classical proofs with operational semantics~\cite{Control1, Parigot1}. In this case we necessarily lose realizability at least of $\exists$ and $\vee$, which - one could argue - is why we want a program in the first place.

We suggest a new approach for obtaining a constructive proof using a classical prover:
\begin{itemize}
	\item Performing our translation on the input formula $\varphi$ to obtain $\varphi^\#$.
	\item Checking the classical validity of $\varphi^\#$ with the prover resulting in a proof or counter-model.
	\item Translating the proof/counter-model of $\varphi^\#$ to a (intuitionistic) proof/counter-model of $\varphi$.
\end{itemize}
